---
title: 'Quality Control'
author: "Ramon Massoni-Badosa"
date: "2021/04/09"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```


# Introduction

We will try to replicate their quality control (QC). Based on the methods section: 

"Low quality cells were excluded from downstream analyses based on %mitochondrial reads <20, features per cell >200 and <2,500, and number of reads per cell <10,000."


# Pre-processing

## Load packages

```{r}
library(Seurat)
library(tidyverse)
library(ggpubr)
library(Matrix)
library(glue)
```


## Define parameters

```{r}
# Paths
path_to_data <- "~/Desktop/richter_transformation/results/R_objects/1.seurat_obj_init_Penter2021.rds"
path_to_save <- "~/Desktop/richter_transformation/results/R_objects/1.seurat_obj_filtered_Penter2021.rds"


# Source functions
# source(here::here("bin/utils.R"))
source("~/Desktop/CNAG/mnt_clust/RICHTER/current/bin/utils.R")


# Parameters
max_n_counts <- 10000
min_n_genes <- 200
max_n_genes <- 2500
max_pct_mt <- 20
n_counts_titl <- "Library Size (total UMI)"
n_genes_titl <- "Number of detected genes"
pct_mt_titl <- "% mitochondrial expression"
n_counts_var <- "nCount_RNA"
n_genes_var <- "nFeature_RNA"
pct_mt_var <- "pct_mt"
```


## Load data

```{r}
seurat <- readRDS(path_to_data)
```


# QC

## Library size (total UMI)

```{r}
plot_histogram_qc(seurat@meta.data, x = n_counts_var, x_lab = n_counts_titl) +
  geom_vline(xintercept = max_n_counts, linetype = "dashed", color = "red")
```


## Number of detected genes

```{r}
plot_histogram_qc(seurat@meta.data, x = n_genes_var, x_lab = n_genes_titl) +
  geom_vline(xintercept = min_n_genes, linetype = "dashed", color = "red") +
  geom_vline(xintercept = max_n_genes, linetype = "dashed", color = "red")
```


## % mitochondrial expression

```{r}
seurat[[pct_mt_var]] <- PercentageFeatureSet(seurat, pattern = "^MT-")
plot_histogram_qc(seurat@meta.data, x = pct_mt_var, x_lab = pct_mt_titl) +
  geom_vline(xintercept = max_pct_mt, linetype = "dashed", color = "red")
```


## Joint

```{r}
# % mitochondrial vs number of detected genes
FeatureScatter(seurat, n_genes_var, pct_mt_var, cols = "black") +
  labs(title = "", x = n_genes_titl, y = pct_mt_titl) +
  NoLegend() +
  geom_vline(xintercept = min_n_genes, linetype = "dashed", color = "red") +
  geom_vline(xintercept = max_n_genes, linetype = "dashed", color = "red") +
  geom_hline(yintercept = max_pct_mt, linetype = "dashed", color = "red")


# number of detected genes vs library size
FeatureScatter(seurat, n_counts_var, n_genes_var, cols = "black") +
  labs(title = "", x = n_counts_titl, y = n_genes_titl) +
  NoLegend() +
  geom_vline(xintercept = max_n_counts, linetype = "dashed", color = "red") +
  geom_hline(yintercept = min_n_genes, linetype = "dashed", color = "red") +
  geom_hline(yintercept = max_n_genes, linetype = "dashed", color = "red")
```


# Filter

```{r}
is_low_quality <- with(seurat@meta.data,
  nCount_RNA > max_n_counts |
  nFeature_RNA < min_n_genes |
  nFeature_RNA > max_n_genes |
  pct_mt > max_pct_mt
)
table(is_low_quality)
seurat <- subset(seurat, cells = colnames(seurat)[!is_low_quality])
```


# Save

```{r}
saveRDS(seurat, path_to_save)
```


# Session information

```{r}
sessionInfo()
```

