---
title: 'Separate microenvironment and tumoral cells'
author: "Ramon Massoni-Badosa"
date: "2021/04/09"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```


# Introduction

Here we will separate the tumoral from the microenvironment cells. To that end, we will perform linear dimensionality reduction (PCA), which we will use to cluster cells.


# Pre-processing

## Load packages

```{r}
library(Seurat)
library(harmony)
library(tidyverse)
library(ggpubr)
library(glue)
```


## Define parameters

```{r}
# Paths
# path_to_data <- "~/Desktop/richter_transformation/results/R_objects/2.seurat_obj_filtered_Penter2021.rds"
# path_to_save <- "~/Desktop/richter_transformation/results/R_objects/3.seurat_obj_clustered_Penter2021.rds"
path_to_data <- here::here("results/R_objects/Penter2021/2.seurat_obj_filtered_Penter2021.rds")
path_to_save <- here::here("results/R_objects/Penter2021/3.seurat_obj_clustered_Penter2021.rds")


# Source functions
source(here::here("bin/utils.R"))
# source("~/Desktop/CNAG/mnt_clust/RICHTER/current/bin/utils.R")


# Colors
color_palette <-  c("#1CFFCE", "#90AD1C", "#C075A6", "#85660D", "#5A5156", "#AA0DFE",   
                    "#F8A19F", "#F7E1A0", "#1C8356", "#FEAF16", "#822E1C", "#C4451C",   
                    "#1CBE4F", "#325A9B", "#F6222E", "#FE00FA", "#FBE426", "#16FF32", 
                    "black",   "#3283FE", "#B00068", "#DEA0FD", "#B10DA1", "#E4E1E3",   
                    "#90AD1C", "#FE00FA", "#85660D", "#3B00FB", "#822E1C", "coral2", 
                    "#1CFFCE", "#1CBE4F", "#3283FE", "#FBE426", "#F7E1A0", "#325A9B",   
                    "#2ED9FF", "#B5EFB5", "#5A5156", "#DEA0FD", "#FEAF16", "#683B79",   
                    "#B10DA1", "#1C7F93", "#F8A19F", "dark orange", "#FEAF16", "#FBE426",  
                    "Brown")
```


## Load data

```{r}
seurat <- readRDS(path_to_data)
```


# Dimensionality reduction

```{r}
seurat <- NormalizeData(seurat)
seurat <- process_seurat(seurat)
DimPlot(
  seurat,
  group.by = "sample_id",
  pt.size = 0.1
) +
  scale_color_manual(values = color_palette)
FeaturePlot(seurat, c("CD3D", "CD79A", "LYZ", "NKG7"))
FeaturePlot(seurat, c("HBB", "TOP2A", "PPBP"))
FeaturePlot(seurat, "TCF4")
```


# Batch effect correction

```{r}
seurat <- RunHarmony(
  seurat,
  group.by.vars = "dataset_id",
  dims = 1:30,
  reduction = "pca"
)
seurat <- RunUMAP(seurat, dims = 1:30, reduction = "harmony")
FeaturePlot(seurat, c("CD3D", "CD79A", "LYZ", "NKG7"))
FeaturePlot(seurat, c("HBB", "TOP2A", "PPBP"))
FeaturePlot(seurat, "TCF4")
```


# Cluster

```{r}
seurat <- FindNeighbors(seurat, reduction = "harmony", dims = 1:30)
seurat <- FindClusters(seurat, resolution = 0.05)
DimPlot(
  seurat,
  group.by = "sample_id",
  pt.size = 0.1
) +
  scale_color_manual(values = color_palette)
FeaturePlot(seurat, c("CD3D", "CD79A", "LYZ", "NKG7"))
FeaturePlot(seurat, c("HBB", "TOP2A", "PPBP"))
FeaturePlot(seurat, "TCF4")
```


# Markers

```{r}
markers_all <- FindAllMarkers(seurat, logfc.threshold = 1, only.pos = TRUE)
DT::datatable(markers_all, options = list(scrollX = TRUE))
```


* Cluster 0: CLL cells
* Cluster 1: CD4 T cells/CD8 T cells/NK
* Cluster 2: myeloid (monocytes/DC)
* Cluster 3: platelets
* Cluster 4: erythrocytes
* Cluster 5: doublets
* Cluster 6: unknown
* Cluster 7: technical
* Cluster 8: unknown


Since the focus of this study is in the CD5+CD19+ B cells, we will subset and save cluster 0:

```{r}
seurat <- subset(seurat, idents = "0")
```


Reprocess

```{r}
seurat <- process_seurat(seurat)
DimPlot(
  seurat,
  group.by = "sample_id",
  pt.size = 0.1
) +
  scale_color_manual(values = color_palette)
```


# Save

```{r}
saveRDS(seurat, path_to_save)
```


# Session information

```{r}
sessionInfo()
```

