---
title: 'Predict Richter cells using a logistic regression model'
author: "Ramon Massoni-Badosa"
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```


# Introduction

Here, we aim to classify unseen single-cell transcriptomes derived from blood samples of CLL patients to predict which are Richter and which are CLL. We will use a [logistic regression](https://en.wikipedia.org/wiki/Logistic_regression) model, following the explanation described in the book ["Introduction to Statistical Learning"](https://www.statlearning.com/) (second edition). We will use the following explanatory variables:

1. OXPHOS score
2. BCR signaling score
3. OXPHOS + BCR score in a multivariate logistic regression model.

The purposes are to assess which variable has more predictive power, and whether using a combination of both variables we can improve our predictions.


## Load packages

```{r packages}
library(Seurat)
library(tidyverse)
library(DescTools)
library(pROC)
```


## Define parameters

```{r}
# Paths
path_to_obj <- here::here("results/R_objects/seurat_training_and_test.rds")


# Source functions
source(here::here("bin/utils.R"))


# Colors
cols_rt <- c("#dcdddc", "#cd8899")
```


## Read data

```{r}
seurat <- readRDS(path_to_obj)
```


# Visualize distribution of explanatory variables


```{r}
# Violin plots + boxplots
training_set <- subset(seurat, is_training == "training")
training_df <- training_set@meta.data[, c("is_rt", "oxphos_score_UCell", "bcr_score_UCell")]
vln_plot_oxphos <- training_df %>%
  ggplot(aes(is_rt, oxphos_score_UCell, fill = is_rt)) +
    geom_violin() +
    geom_boxplot(width = 0.25) +
    scale_fill_manual(values = cols_rt) +
    ylab("OXPHOS score") +
    theme_classic() +
    theme(legend.position = "none", axis.title.x = element_blank())
vln_plot_bcr <- training_df %>%
  ggplot(aes(is_rt, bcr_score_UCell, fill = is_rt)) +
    geom_violin() +
    geom_boxplot(width = 0.25) +
    scale_fill_manual(values = cols_rt) +
    ylab("BCR signaling score") +
    theme_classic() +
    theme(legend.position = "none", axis.title.x = element_blank())
vln_plot_oxphos | vln_plot_bcr


# Scatter plot
scatter_plot <- training_df %>%
  ggplot(aes(oxphos_score_UCell, bcr_score_UCell, color = is_rt)) +
    geom_point(size = 0.25) +
    scale_color_manual(values = cols_rt) +
    labs(x = "OXPHOS score", y = "BCR signaling score") +
    theme_classic() +
    theme(legend.position = "none")
scatter_plot
```


# Train models

```{r}
training_df$is_rt <- ifelse(training_df$is_rt == "RT", 1, 0)
oxphos_fit <- glm(is_rt ~ oxphos_score_UCell, data = training_df, family = "binomial")
bcr_fit <- glm(is_rt ~ bcr_score_UCell, data = training_df, family = "binomial")
oxphos_bcr_fit <- glm(is_rt ~ ., data = training_df, family = "binomial")
summary(oxphos_fit)
summary(bcr_fit)
summary(oxphos_bcr_fit)
```



# Test models

```{r}
# Predict probabilities on test set
test_set <- subset(seurat, is_training != "training")
test_df <- test_set@meta.data[, c("is_rt", "oxphos_score_UCell", "bcr_score_UCell")]
test_df$is_rt <- ifelse(test_df$is_rt == "RT", 1, 0)
test_df$probabilities_oxhpos <- predict(oxphos_fit,newdata = test_df, type = "response")
test_df$probabilities_bcr <- predict(bcr_fit, newdata = test_df, type = "response")
test_df$probabilities_oxhpos_bcr <- predict(oxphos_bcr_fit, newdata = test_df, type = "response")


# Plot sigmoidal curves
sigm_oxphos <- test_df %>%
  mutate(is_rt = ifelse(is_rt == 1, "RT", "CLL")) %>% 
  ggplot(aes(oxphos_score_UCell, probabilities_oxhpos, color = is_rt)) +
    geom_point(size = 0.75) +
    scale_color_manual(values = cols_rt) +
    labs(x = "OXPHOS Score", y = "P(RT)") +
    theme_classic() +
    theme(legend.title = element_blank()) +
    guides(color = guide_legend(override.aes = list(size = 2)))
sigm_oxphos
sigm_bcr <- test_df %>%
  mutate(is_rt = ifelse(is_rt == 1, "RT", "CLL")) %>% 
  ggplot(aes(bcr_score_UCell, probabilities_bcr, color = is_rt)) +
    geom_point(size = 0.75) +
    scale_color_manual(values = cols_rt) +
    labs(x = "BCR signaling score", y = "P(RT)") +
    theme_classic() +
    theme(legend.title = element_blank()) +
    guides(color = guide_legend(override.aes = list(size = 2)))
sigm_bcr


# Area under the curve (AUC)
roc_oxphos <- pROC::roc(
  response = test_df$is_rt,
  predictor = test_df$probabilities_oxhpos
)
roc_bcr <- pROC::roc(
  response = test_df$is_rt,
  predictor = test_df$probabilities_bcr
)
roc_oxphos_bcr <- pROC::roc(
  response = test_df$is_rt,
  predictor = test_df$probabilities_oxhpos_bcr
)
plot(roc_oxphos)
plot(roc_bcr)
plot(roc_oxphos_bcr)
roc_oxphos$auc[1]
roc_bcr$auc[1]
roc_oxphos_bcr$auc[1]
```


# Session Information

```{r}
sessionInfo()
```

