---
title: 'Separate leukemic and microenvironment cells'
author: "Ramon Massoni-Badosa"
date: "2021/04/09"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```


# Introduction

The objective of this notebook is to separate the microenvironment (T cells, monocytes, etc.) from the leukemic B (CD79+) cells.


# Pre-processing

## Load packages

```{r}
library(Seurat)
library(SeuratWrappers)
library(harmony)
library(tidyverse)
```


## Define parameters

```{r}
# Paths
path_to_obj <- here::here("results/R_objects/3.seurat_filtered.rds")
path_to_save_microenv <- here::here("results/R_objects/4.seurat_microenvironment.rds")
path_to_save_leukemic <- here::here("results/R_objects/4.seurat_leukemic.rds")


# Colors
color_palette <- c("black", "gray", "red", "yellow", "violet", "green4",
                   "blue", "mediumorchid2", "coral2", "blueviolet",
                   "indianred4", "deepskyblue1", "dimgray", "deeppink1",
                   "green", "lightgray", "hotpink1")


# Source functions
source(here::here("bin/utils.R"))
```


## Load data

```{r}
seurat <- readRDS(path_to_obj)
```


# Normalization, Dimensionality reduction and batch effect correction

```{r}
seurat <- seurat %>%
  NormalizeData(
    normalization.method = "LogNormalize",
    scale.factor = 10000
  ) %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  RunHarmony(group.by.vars = "sample_id", reduction = "pca", dims = 1:30) %>%
  RunUMAP(dims = 1:30, reduction = "harmony")
```


# Visualize lineage markers

To ease the annotation, let us visualize some key lineage markers

```{r}
lineage_markers <- c("CD79A", "CD3D", "CD8A", "IL7R", "NKG7",
                     "LYZ", "HBM", "TOP2A")
lineage_markers_ggs <- purrr::map(lineage_markers, function(x) {
  p <- FeaturePlot(seurat, features = x, pt.size = 0.45)
  p +
    scale_color_viridis_c(option = "magma")
})
names(lineage_markers_ggs) <- lineage_markers
lineage_markers_ggs
```


# Cluster

```{r}
seurat <- FindNeighbors(
  seurat,
  dims = 1:30,
  reduction = "harmony"
)
seurat <- FindClusters(seurat, resolution = 0.1)
DimPlot(seurat, cols = color_palette, pt.size = 0.3)
richter_levels <- c("richter_sp", "richter_while-ibrutinib",
                    "richter_post-ofatu-duvelisib")
seurat$is_richter <- ifelse(
  seurat$sample_description %in% richter_levels,
  TRUE,
  FALSE
)
DimPlot(seurat, group.by = "is_richter")
DimPlot(seurat, group.by = "tissue")
```


In addition, we have seen a cluster of erythroblasts (HBM+) that cluster with the CD79A+ cells, most likely due to its strong proliferative signature. We will fetch them with the `FindSubCluster` function and rule them out from the analysis:

```{r}
# Subcluster
seurat <- FindSubCluster(
  seurat,
  cluster = "3",
  graph.name = "RNA_snn",
  subcluster.name = "fetch_erythroblasts",
  resolution = 0.2
)
DimPlot(seurat, group.by = "fetch_erythroblasts", cols = color_palette)
Idents(seurat) <- "fetch_erythroblasts"


# Find markers of potential erythroblasts
markers_erythroblasts <- purrr::map(c("3_3", "3_4"), function(x) {
  df <- FindMarkers(
    seurat,
    ident.1 = x,
    only.pos = TRUE,
    logfc.threshold = 0.75
  )
  df <- df %>%
    rownames_to_column("gene") %>%
    arrange(desc(avg_log2FC))
  df
})
names(markers_erythroblasts) <- c("3_3", "3_4")
DT::datatable(markers_erythroblasts$`3_3`)
DT::datatable(markers_erythroblasts$`3_4`)


# Subset
mask <- colnames(seurat)[!(seurat$fetch_erythroblasts %in% c("3_3", "3_4"))]
seurat <- subset(seurat, cells = mask)
DimPlot(seurat, cols = color_palette)
```


# Separate CD5+ B cells from microenvironment

As we can see, clusters 2 and 4 correspond to T cells, NK cells and monocytes.

```{r}
microenv_clusters <- c("2", "4")
leuk_clusters <- c("0", "1", "3_0", "3_1", "3_2")
microenv <- subset(seurat, idents = microenv_clusters)
leuk <- subset(seurat, idents = leuk_clusters)
DimPlot(leuk, cols = color_palette, pt.size = 0.3)
DimPlot(microenv, cols = color_palette, pt.size = 0.3)
```


# Save

```{r}
saveRDS(microenv, path_to_save_microenv)
saveRDS(leuk, path_to_save_leukemic)
```


# Sesion Information

```{r}
sessionInfo()
```

