---
title: 'Cluster tumoral cells (case 3299)'
author: "Ramon Massoni-Badosa"
date: "2021/04/09"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```


# Introduction

Here, we will cluster the cells of the patient with id "3299".


# Pre-processing

## Load packages

```{r}
library(Seurat)
library(SeuratWrappers)
library(harmony)
library(tidyverse)
```


## Define parameters

```{r}
# Paths
path_to_obj <- here::here("results/R_objects/4.seurat_leukemic.rds")
path_to_save <- here::here("results/R_objects/5.seurat_clustered_3299.rds")


# Colors
color_palette <- c("black", "gray", "red", "yellow", "violet", "green4",
                   "blue", "mediumorchid2", "coral2", "blueviolet",
                   "indianred4", "deepskyblue1", "dimgray", "deeppink1",
                   "green", "lightgray", "hotpink1")


# Source functions
source(here::here("bin/utils.R"))


# Thresholds
```

  
## Load data

```{r}
seurat <- readRDS(path_to_obj)
```


# Subset and reprocess

Of note, everytime that we filter out a specific population, the set of highly variable genes (HVG) and the axis of variability (PCs) changes. Thus, we need to rerun the main analysis steps iteratively.

First, we focus on the patient studied in this notebook:

```{r}
seurat <- subset(seurat, donor_id == "3299")
seurat <- process_seurat(seurat, dims = 1:20)
DimPlot(seurat, group.by = "subproject")
```

In this case we only have one subproject, so we do not need to worry about batch effects. However, we see that the Richter transformation (RT) sample comes from peripheral blood (PB) whilst the other come from bone marrow (BM)

```{r}
table(seurat$sample_description_FN, seurat$tissue)
DimPlot(seurat, group.by = "tissue")
```


Thus, we will focus on the samples that come from BM. Since we know that Ferran could assign 80 to 90% of the cells in post-treatment_3 to Richter cells, we will work with them and exclude the RT sample:

```{r}
seurat <- subset(seurat, tissue == "BM")
seurat <- process_seurat(seurat, dims = 1:20)
DimPlot(seurat, group.by = "tissue")
DimPlot(seurat, group.by = "time_point")
FeaturePlot(seurat, "pct_mt") + scale_color_viridis_c(option = "magma")
FeaturePlot(seurat, "nFeature_RNA") + scale_color_viridis_c(option = "magma")
FeaturePlot(seurat, "HBM") + scale_color_viridis_c(option = "magma")
```


We see a subpopulation with characteristics of low-quality cells: high mitochondrial expression and low number of detected features (genes). In addition, we still see a lingering cluster of erythroblasts (HBM+). Hence, let us exclude both from the dataset:

```{r}
seurat <- FindNeighbors(seurat, reduction = "pca", dims = 1:20)
seurat <- FindClusters(seurat, resolution = 0.75)
DimPlot(seurat)
selected_cells <- colnames(seurat)[!(seurat$seurat_clusters %in% c("7", "8"))]
seurat <- subset(seurat, cells = selected_cells)
seurat <- process_seurat(seurat, dims = 1:20)
DimPlot(seurat)
DimPlot(seurat, group.by = "time_point")
```


# Cluster

```{r}
seurat <- FindNeighbors(seurat, reduction = "pca", dims = 1:20)
seurat <- FindClusters(seurat, resolution = 0.3)
DimPlot(seurat)
```


Cluster 2 can be further stratified:

```{r}
seurat <- FindSubCluster(
  seurat,
  cluster = "2",
  graph.name = "RNA_snn",
  subcluster.name = "subcluster_2",
  resolution = 0.4
)
DimPlot(seurat, group.by = "subcluster_2")
FeaturePlot(seurat, c("CD27", "S100A4", "HBD"))
```


Despite our efforts, we see that we still have some lingering erythroblasts (HBD+). Let us remove them:

```{r}
seurat <- subset(seurat, subcluster_2 != "3")
seurat <- process_seurat(seurat, dims = 1:20)
DimPlot(seurat, group.by = "subcluster_2")
```

Finally, in later stages of the project we noticed that we had lingering doublets with cells from the microenvironment (T cells, NK cells or monocytes):

```{r}
FeaturePlot(seurat, c("CD3D", "NKG7", "LYZ"))
```

```{r}
seurat2 <- seurat
seurat2 <- FindNeighbors(seurat2, k.param = 5, reduction = "pca", dims = 1:20)
seurat2 <- FindClusters(seurat2, resolution = 0.2)
DimPlot(seurat2)
selected_cells <- colnames(seurat2)[seurat2$seurat_clusters != "3"]
seurat <- subset(seurat, cells = selected_cells)
DimPlot(seurat, group.by = "subcluster_2")
```


Finally, we observe that, despite some differences, clusters 2_0 and 2_1 both express CD27 and S100A4. To keep the consistency with previous annotations, we will group them into a single cluster:

```{r}
seurat$final_clusters <- factor(case_when(
  seurat$subcluster_2 == "0" ~ "0",
  seurat$subcluster_2 == "1" ~ "1",
  seurat$subcluster_2 == "2_0" ~ "2",
  seurat$subcluster_2 == "2_1" ~ "2",
  seurat$subcluster_2 == "2_2" ~ "3",
  seurat$subcluster_2 == "2_3" ~ "4", 
))
Idents(seurat) <- "final_clusters"
DimPlot(seurat)
```


# Save

```{r}
saveRDS(seurat, path_to_save)
```


# Session Info

```{r}
sessionInfo()
```

