---
title: 'Annotate tumoral cells (case 12)'
author: "Ramon Massoni-Badosa"
date: "2021/04/09"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```


# Introduction

Here, we will annotate the cells of the patient with id "012".


# Pre-processing

## Load packages

```{r}
library(Seurat)
library(SeuratWrappers)
library(harmony)
library(tidyverse)
```


## Define parameters

```{r}
# Paths
path_to_obj <- here::here("results/R_objects/5.seurat_clustered_12.rds")
path_to_save <- here::here("results/R_objects/6.seurat_annotated_12.rds")
path_to_save_markers <- here::here("3-clustering_and_annotation/tmp/markers_clusters_12.rds")


# Colors
color_palette <- c("black", "gray", "red", "yellow", "violet", "green4",
                   "blue", "mediumorchid2", "coral2", "blueviolet",
                   "indianred4", "deepskyblue1", "dimgray", "deeppink1",
                   "green", "lightgray", "hotpink1")


# Source functions
source(here::here("bin/utils.R"))


# Thresholds
min_log2FC <- 0.3
alpha <- 0.001
```

  
## Load data

```{r}
seurat <- readRDS(path_to_obj)
seurat
DimPlot(seurat, cols = color_palette)
```


# Cell Cycle Score

As upregulation of cell cycle genes is a hallmark of Richter transformation, we will infer the cell cycle score and phase for each cell:

```{r}
seurat <- CellCycleScoring(
  seurat,
  s.features = cc.genes.updated.2019$s.genes,
  g2m.features = cc.genes.updated.2019$g2m.genes,
  set.ident = FALSE
)
DimPlot(seurat, group.by = "Phase")
umap_s_score <- FeaturePlot(seurat, features = "S.Score") +
  scale_color_viridis_c(option = "magma") +
  labs(title = "S Score") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "plain"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank()
  )
umap_g2m_score <- FeaturePlot(seurat, features = "G2M.Score") +
  scale_color_viridis_c(option = "magma") +
  labs(title = "G2M Score") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "plain"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank()
  )
umap_cc_combined <- ggpubr::ggarrange(
  plotlist = list(umap_s_score, umap_g2m_score),
  nrow = 2,
  ncol = 1,
  common.legend = FALSE
)
umap_cc_combined
```


# Find Markers

```{r}
markers <- FindAllMarkers(seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)
DT::datatable(markers, options = list(scrollX = TRUE))
```


# Annotation

Important literature to annotate the cells:


* [Intraclonal Complexity in Chronic Lymphocytic Leukemia:Fractions Enriched in Recently Born/Divided andOlder/Quiescent Cells](https://molmed.biomedcentral.com/articles/10.2119/molmed.2011.00360).
* [MicroRNA-155influences B-cell receptor signaling and associateswith aggressive disease in chronic lymphocytic leukemia](https://ashpublications.org/blood/article/124/4/546/33343/MicroRNA-155-influences-B-cell-receptor-signaling).


| Cluster | Markers                   | Annotation                 |
|---------|---------------------------|----------------------------|
| 0       | CXCR4                     | CXCR4hiCD27lo              |
| 1       | CD27                      | CXCR4loCD27hi              |
| 2_0     | CCND2                     | CCND2hi RT                 |
| 2_1     | PAGE2, PAGE2B             | CCND2lo RT                 |
| 3       | TOP2A, MKI67              | RT proliferative           |
| 4       | MIR155HG                  | MIR155HGhi                 |
| 5       | IGHM, MZB1, XBP1          | MZB1hiIGHMhiXBP1hi         |


```{r}
seurat$annotation_final <- factor(
  seurat$final_clusters,
  levels = c("0", "1", "2_0", "2_1", "3", "4", "5"),
)
new_levels_12 <- c("CXCR4hiCD27lo", "CXCR4loCD27hi", "CCND2hi RT",
                   "CCND2lo RT", "RT proliferative", "MIR155HGhi",
                   "MZB1hiIGHMhiXBP1hi")
levels(seurat$annotation_final) <- new_levels_12
seurat$annotation_final <- factor(
  seurat$annotation_final,
  color_annotations$`12`$annotation_final
)
Idents(seurat) <- "annotation_final"


# Plot UMAP
(umap_annotation <- plot_annotation(
  seurat_obj = seurat,
  pt_size = 0.5,
  colors_reference = color_annotations,
  patient_id = "12",
  nothing = TRUE
))
```


# Visualize markers

```{r}
# UMAPs
genes_interest <- c("CXCR4", "CD24", "CD27", "MIR155HG", "CCND2", "TOP2A",
                    "PCNA", "MZB1", "IGHM", "XBP1")
(feature_plots <- purrr::map(genes_interest, function(x) {
  p <- FeaturePlot(seurat, x, pt.size = 0.5) +
    scale_color_viridis_c(option = "magma")
  p
}))


# Dot plots
(dot_plot <- plot_dot_plot(
  seurat,
  goi = rev(genes_interest),
  colors_reference = color_annotations,
  patient_id = "12"
))


# Violin plots
(vln_plot_s <- plot_violin_plot(
  seurat,
  continuous_var = "S.Score",
  ylab = "S Phase Score",
  colors_reference = color_annotations,
  patient_id = "12"
))
(vln_plot_g2m <- plot_violin_plot(
  seurat,
  continuous_var = "G2M.Score",
  ylab = "G2M Phase Score",
  colors_reference = color_annotations,
  patient_id = "12"
))
```


# Save

```{r}
# Save Seurat object
saveRDS(seurat, path_to_save)


# Save markers
markers$annotation <- factor(markers$cluster)
levels(markers$annotation) <- new_levels_12
markers_list <- purrr::map(levels(markers$annotation), function(x) {
  df <- markers[markers$annotation == x, ]
  df <- df[, c(7, 1, 5, 2:4, 6, 8)]
  df
})
names(markers_list) <- levels(markers$annotation)
markers_list <- markers_list[color_annotations$`12`$annotation_final]
markers_final <- bind_rows(markers_list)
saveRDS(markers_list, path_to_save_markers)
saveRDS(
  markers_final,
  here::here("results/tables/markers/markers_annotated_clusters_patient_12.rds")
)
openxlsx::write.xlsx(
  x = markers_list,
  file = here::here("results/tables/markers/markers_annotated_clusters_patient_12.xlsx")
)
```


# Session Info

```{r}
sessionInfo()
```