---
title: 'Time-point-specific clustering'
author: "Ramon Massoni-Badosa"
date: "2021/06/07"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```


# Introduction

Here, we will perform patient- and time-point-specific clustering. If we can indeed predict Richter-like cells at the early stages of the disease, then they should cluster together without having the Richter time-point present. On the other hand, if they were poor-quality cells, then they might scatter everywhere in the absence of a Richter-like cluster.


# Pre-processing

## Load packages

```{r}
library(Seurat)
library(tidyverse)
```


## Define parameters

```{r}
# Paths
path_to_12 <- here::here("results/R_objects/6.seurat_annotated_12.rds")
path_to_19 <- here::here("results/R_objects/6.seurat_annotated_19.rds")
path_to_63 <- here::here("results/R_objects/patient_63/3.seurat_annotated.rds")
path_to_365 <- here::here("results/R_objects/6.seurat_annotated_365.rds")
path_to_3299 <- here::here("results/R_objects/6.seurat_annotated_3299.rds")
path_to_save <- here::here("results/R_objects/7.seurat_time_point_specific_list.rds")


# Source functions
source(here::here("bin/utils.R"))
```


## Load data

```{r}
paths_to_load <- c(
  "12" = path_to_12,
  "19" = path_to_19,
  "63" = path_to_63,
  "365" = path_to_365,
  "3299" = path_to_3299
)
seurat_list <- purrr::map(paths_to_load, readRDS)
seurat_list
umaps_annotations <- purrr::map(names(seurat_list), function(x) {
  p <- plot_annotation(
    seurat_list[[x]],
    pt_size = 0.5,
    colors_reference = color_annotations,
    patient_id = x
  )
  p +
    ggtitle(x) +
    theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
})
umaps_annotations
```


# Time point-specific dimensionality reduction

```{r}
seurat_list_list <- purrr::map(seurat_list, function(seurat_obj) {
  seurat_obj$time_point2 <- str_c(
    seurat_obj$time_point,
    seurat_obj$tissue,
    sep = "_"
  )
  seurat_obj_list <- SplitObject(seurat_obj, split.by = "time_point2")
  seurat_obj_list <- seurat_obj_list[sort(unique(seurat_obj$time_point2))]
  seurat_obj_list <- purrr::map(seurat_obj_list, process_seurat, dims = 1:20)
  seurat_obj_list
})
```


# Plot

```{r}
umaps_time_points <- purrr::map2(seurat_list_list, names(seurat_list_list), function(seurat_obj_list, patient_id) {
  purrr::map2(seurat_obj_list, names(seurat_obj_list), function(seurat_obj, x) {
    p <- plot_annotation(
      seurat_obj = seurat_obj,
      pt_size = 0.65,
      colors_reference = color_annotations,
      patient_id = patient_id,
      nothing = TRUE
    )
    p +
      ggtitle(x) +
      theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
  })
})

print("Case 12")
umaps_time_points$`12`

print("Case 19")
umaps_time_points$`19`

print("Case 63")
umaps_time_points$`63`

print("Case 365")
umaps_time_points$`365`

print("Case 3299")
umaps_time_points$`3299`
```


# Save

```{r}
saveRDS(seurat_list_list, path_to_save)
```



# Session Information

```{r}
sessionInfo()
```

