---
title: 'Technical replicate (BCLLATLAS_29)'
author: "Ramon Massoni-Badosa"
date: "2021/06/07"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```


# Introduction

Here, we will analyze the technical replicate of specific clinical samples. We will use them as technical replicates to double-check that we can indeed find RT-like cells in early points of the disease.


# Pre-processing

## Load packages

```{r}
library(Seurat)
library(ggpubr)
library(harmony)
library(symphony)
library(tidyverse)
```


## Define parameters

```{r}
path_to_obj <- here::here("results/R_objects/4.seurat_leukemic.rds")
path_to_reference_12 <- here::here("results/R_objects/6.seurat_annotated_12.rds")
path_to_dea <- here::here("6-differential_expression_analysis/tmp/patient_specific_differential_expression_analysis_rt_vs_cll.rds")


# Source functions and variables
source(here::here("bin/utils.R"))
```


## Load data

```{r}
seurat <- readRDS(path_to_obj)
seurat_12_ref <- readRDS(path_to_reference_12)
dea <- readRDS(path_to_dea)
```


# Exploratory data analysis

```{r}
# Split objects by donor and time-point
seurat <- subset(seurat, subproject == "BCLLATLAS_29")
seurat_list <- SplitObject(seurat, "donor_id")
seurat_list <- seurat_list[c("12", "19")]
seurat_list$`12` <- SplitObject(seurat_list$`12`, "time_point")
seurat_list$`19` <- SplitObject(seurat_list$`19`, "time_point")


# Dimensionality reduction
seurat_list$`12` <- purrr::map(seurat_list$`12`, process_seurat, dims = 1:20)
seurat_list$`19`$T1 <- process_seurat(seurat_list$`19`$T1, dims = 1:20)


# Visualize
DimPlot(seurat_list$`12`$T1)
DimPlot(seurat_list$`12`$T2)
DimPlot(seurat_list$`19`$T1)
```


Judging by our previous QC and the UMAPs above, we will focus on the T1 of case 12, since it is the only one for which we have enough cells and detected genes, and we can unravel heterogeneity.

Calculate signatures:

```{r}
# Minimal QC
seurat <- seurat_list$`12`$T1
FeaturePlot(seurat, "pct_mt")
seurat <- subset(seurat, pct_mt < 9)
seurat <- subset(seurat, nFeature_RNA > 750)
seurat <- process_seurat(seurat, dims = 1:20)
DimPlot(seurat)
FeaturePlot(seurat, "MIR155HG", order = TRUE, pt.size = 1) + scale_color_viridis_c(option = "magma")


# Calculate signatures
dea <- dea$`12`
dea <- arrange(dea, avg_log2FC)
cll_signature <- dea$gene[1:250]
dea <- arrange(dea, desc(avg_log2FC))
rt_signature <- dea$gene[1:250]
seurat <- AddModuleScore(
  seurat,
  features = list(
    cll_signature = cll_signature,
    rt_signature = rt_signature
  ),
  name = c("cll_signature", "rt_signature")
)
FeaturePlot(seurat, c("cll_signature1", "rt_signature2"), order = TRUE, pt.size = 0.7) &
  scale_color_viridis_c(option = "magma")
FeatureScatter(
  seurat,
  feature1 = "cll_signature1",
  feature2 = "rt_signature2",
  pt.size = 0.7
)
```

