---
title: 'Clustering and annotation (patient 63)'
author: "Ramon Massoni-Badosa"
date: "2021/04/09"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```


# Introduction

The purpose of this notebook is to cluster and annotate the cells obtained from patient with id 63.


# Pre-processing

## Load packages

```{r}
library(Seurat)
library(ggpubr)
library(tidyverse)
```


## Parameters

```{r}
# Paths
path_to_obj <- here::here("results/R_objects/patient_63/2.seurat_object_filtered.rds")
path_to_save <- here::here("results/R_objects/patient_63/3.seurat_annotated.rds")
path_to_save_markers <- here::here("3-clustering_and_annotation/tmp/markers_clusters_63.rds")


# Functions
source(here::here("bin/utils.R"))


# Params
k_param <- 10
min_log2FC <- 0.3
alpha <- 0.001
```


## Load data

```{r}
seurat <- readRDS(path_to_obj)
seurat
```


# Dimensionality reduction

## All cells

### Data normalization

```{r}
seurat <- NormalizeData(
  seurat,
  normalization.method = "LogNormalize",
  scale.factor = 10000
)
```


### Highly variable genes

```{r}
seurat <- FindVariableFeatures(seurat)
LabelPoints(
  plot = VariableFeaturePlot(seurat),
  points = head(VariableFeatures(seurat), 10),
  repel = TRUE
)
```

### Principal Component Analysis

```{r}
seurat <- ScaleData(seurat)
seurat <- RunPCA(seurat)
VizDimLoadings(seurat, dims = 1:2, reduction = "pca")
```


### UMAP

```{r fig.wide=TRUE}
seurat <- RunUMAP(seurat, reduction = "pca", dims = 1:20)
umap_time_point <- DimPlot(seurat, group.by = "time_point")
umap_tissue <- DimPlot(seurat, group.by = "tissue")
umap_time_point + umap_tissue
```

# Cell Cycle Score

As upregulation of cell cycle genes is a hallmark of Richter transformation, we will infer the cell cycle score and phase for each cell:

```{r}
seurat <- CellCycleScoring(
  seurat,
  s.features = cc.genes.updated.2019$s.genes,
  g2m.features = cc.genes.updated.2019$g2m.genes,
  set.ident = FALSE
)
DimPlot(seurat, group.by = "Phase")
umap_s_score <- FeaturePlot(seurat, features = "S.Score") +
  scale_color_viridis_c(option = "magma") +
  labs(title = "S Score") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "plain"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank()
  )
umap_g2m_score <- FeaturePlot(seurat, features = "G2M.Score") +
  scale_color_viridis_c(option = "magma") +
  labs(title = "G2M Score") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "plain"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank()
  )
umap_cc_combined <- ggpubr::ggarrange(
  plotlist = list(umap_s_score, umap_g2m_score),
  nrow = 2,
  ncol = 1,
  common.legend = FALSE
)
umap_cc_combined
```


# Cluster

```{r}
seurat <- FindNeighbors(
  seurat,
  k.param = k_param,
  dims = 1:20,
  reduction = "pca"
)
seurat <- FindClusters(seurat, resolution = 0.2)
DimPlot(seurat)
```

Let us subcluster to find the subpopulation of cycling cells

```{r}
seurat <- FindSubCluster(
  seurat,
  cluster = "1",
  graph.name = "RNA_snn",
  subcluster.name = "subcluster_proliferative",
  resolution = 0.3
)
DimPlot(seurat, group.by = "subcluster_proliferative")
```

# Markers

```{r}
seurat$final_clusters <- seurat$subcluster_proliferative
Idents(seurat) <- "final_clusters"
markers <- FindAllMarkers(seurat, only.pos = TRUE, logfc.threshold = min_log2FC)
markers <- markers %>%
  mutate(cluster = as.character(cluster)) %>%
  dplyr::filter(p_val_adj < alpha) %>%
  arrange(cluster) %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE)
DT::datatable(markers, options = list(scrollX = TRUE))
```


# Annotation

| Cluster | Markers                   | Annotation                 |
|---------|---------------------------|----------------------------|
| 0       | CXCR4                     | CLL                        |
| 1_0     | TCL1A, BTK, WNT3          | RT quiescent               |
| 1_1     | TOP2A, PCNA               | RT proliferative           |


```{r}
seurat$annotation_final <- factor(
  seurat$final_clusters,
  levels = c("0", "1_0", "1_1"),
)
new_levels_63 <- c("CLL", "RT quiescent", "RT proliferative")
levels(seurat$annotation_final) <- new_levels_63
seurat$annotation_final <- factor(
  seurat$annotation_final,
  color_annotations$`63`$annotation_final
)
Idents(seurat) <- "annotation_final"


# Plot UMAP
(umap_annotation <- plot_annotation(
  seurat_obj = seurat,
  pt_size = 0.5,
  colors_reference = color_annotations,
  patient_id = "63",
  nothing = TRUE
))
```


# Visualize markers

```{r}
# UMAPs
genes_interest <- c("CXCR4", "TCL1A", "BTK", "WNT3", "TOP2A", "PCNA")
(feature_plots <- purrr::map(genes_interest, function(x) {
  p <- FeaturePlot(seurat, x, pt.size = 1) +
    scale_color_viridis_c(option = "magma")
  p
}))


# Dot plots
(dot_plot <- plot_dot_plot(
  seurat,
  goi = rev(genes_interest),
  colors_reference = color_annotations,
  patient_id = "63"
))


# Violin plots
(vln_plot_s <- plot_violin_plot(
  seurat,
  continuous_var = "S.Score",
  ylab = "S Phase Score",
  colors_reference = color_annotations,
  patient_id = "63"
))
(vln_plot_g2m <- plot_violin_plot(
  seurat,
  continuous_var = "G2M.Score",
  ylab = "G2M Phase Score",
  colors_reference = color_annotations,
  patient_id = "63"
))
```


# Save

```{r}
# Save Seurat object
saveRDS(seurat, path_to_save)


# Save markers
markers$annotation <- factor(markers$cluster)
levels(markers$annotation) <- new_levels_63
markers_list <- purrr::map(levels(markers$annotation), function(x) {
  df <- markers[markers$annotation == x, ]
  df <- df[, c(7, 1, 5, 2:4, 6, 8)]
  df
})
names(markers_list) <- levels(markers$annotation)
markers_list <- markers_list[color_annotations$`63`$annotation_final]
markers_final <- bind_rows(markers_list)
saveRDS(markers_list, path_to_save_markers)
saveRDS(
  markers_final,
  here::here("results/tables/markers/markers_annotated_clusters_patient_63.rds")
)
openxlsx::write.xlsx(
  x = markers_list,
  file = here::here("results/tables/markers/markers_annotated_clusters_patient_63.xlsx")
)
```


# Session Information

```{r}
sessionInfo()
```


