---
title: 'Quality Control: Cell and Gene Filtering'
author: "Ramon Massoni-Badosa"
date: "2021/04/09"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```


# Introduction

The purpose of this notebook is to filter out poor quality cells and genes.


# Pre-processing

## Load packages

```{r}
library(Seurat)
library(ggpubr)
library(tidyverse)
```


## Parameters

```{r}
# Paths
path_to_obj <- here::here("results/R_objects/2.seurat_unfiltered.rds")
path_to_project_metadata <- here::here("1-cellranger_mapping/data/richter_metadata.csv")
path_to_sample_metadata <- here::here("data/sample_metadata.csv")
path_to_save <- here::here("results/R_objects/3.seurat_filtered.rds")


# Functions
source(here::here("bin/utils.R"))


# Thresholds
min_lib_size <- 900
min_n_genes <- 200
max_pct_mt <- 30
min_cells <- 3
```


## Load data

```{r}
# Read merged Seurat object
seurat <- readRDS(path_to_obj)
seurat


# Read project and donor metadata
project_metadata_df <- read_csv(path_to_project_metadata)
DT::datatable(project_metadata_df)
sample_metadata_df <- read_csv(path_to_sample_metadata)
DT::datatable(sample_metadata_df)
```


## Add sample metadata

```{r}
sample_metadata_df$donor_id <- as.character(sample_metadata_df$donor_id)
seurat$donor_id <- seurat$donor_id %>%
  str_remove("ICGC_") %>%
  str_remove("^0")
seurat$subproject <- ifelse(seurat$is_hashed, "BCLLATLAS_10", "BCLLATLAS_29")
seurat$sample_id <- seurat$hash.ID
singlets <- colnames(seurat)[!(seurat$hash.ID %in% c("Negative", "Doublet"))]
seurat <- subset(seurat, cells = singlets)
seurat$sample_id <- seurat$sample_id %>%
  str_remove("(ICGC-012-|ICGC-019|ICGC-365-|ICGC-3299-)") %>%
  str_remove("^-")
seurat$sample_id[seurat$sample_id == "NA"] <- seurat$library_name[seurat$sample_id == "NA"] 
seurat$sample_id <- str_replace(seurat$sample_id, "(-|_)", "/")
seurat$sample_id[seurat$sample_id == "365/07"] <- "365-07"
seurat$sample_id[seurat$sample_id == "19/06"] <- "19-06"
seurat$sample_id[seurat$sample_id == "19/12"] <- "19-12"
sample_metadata_df[sample_metadata_df$sample_id == "19/2768B", ]


test_identifiers <- all(unique(sample_metadata_df$sample_id) %in% unique(sample_metadata_df$sample_id))
if (test_identifiers) {
  sample_metadata_df <- select(sample_metadata_df, - "donor_id")
  new_metadata <- seurat@meta.data %>%
    left_join(sample_metadata_df, by = "sample_id") %>%
    as.data.frame()
  rownames(new_metadata) <- colnames(seurat)
  seurat@meta.data <- new_metadata

} else {
  stop("Identifiers are different!")
}


# Remove hashing-specific columns
removed_vars <- c("HTO_maxID", "HTO_secondID", "HTO_margin", "HTO_classification",
                  "HTO_classification.global")
seurat@meta.data <- seurat@meta.data[, !(colnames(seurat@meta.data) %in% removed_vars)]
```


# Distribution of QC variables

First, we want to see how the main QC variables distribute across samples. This will help us interpret the results downstream, as we will know which are the samples with lower quality.


## Signal-to-noise ratio (cell hashing)

```{r}
# Calculate mean SNR
snr_means_df <- seurat@meta.data %>%
  dplyr::filter(is_hashed) %>%
  dplyr::group_by(library_name) %>%
  dplyr::summarise(
    mean_snr = round(mean(hashing_snr), 2),
    sd_snr = round(sd(hashing_snr), 2)
  ) %>%
  dplyr::arrange(desc(mean_snr))
DT::datatable(snr_means_df)


# Plot SNR distribution
breaks <- c(0, 2.5, 5, 7.5)
snr_gg <- seurat@meta.data %>%
  dplyr::filter(is_hashed) %>%
  horizontal_boxplot(
    categorical_var = "library_name",
    continuous_var = "hashing_snr",
    fill = "donor_id",
    ylab = "Signal-to-Noise (hashing)"
  ) +
  scale_y_continuous(limits = c(0, 7.5), breaks = breaks)
snr_gg
```


## Library complexity

```{r}
# Calculate mean library size
seurat_list <- SplitObject(seurat, split.by = "donor_id")
n_features_means_dfs <- purrr::map(seurat_list, function(seurat_obj) {
  df <- seurat_obj@meta.data %>%
    dplyr::group_by(time_point, sample_id) %>%
    dplyr::summarise(
      mean_n_features = round(mean(nFeature_RNA), 2),
      sd_n_features = round(sd(nFeature_RNA), 2)) %>%
    dplyr::arrange(time_point)
  df
})
n_features_means_dfs


# Plot library size distribution
n_features_ggs <- purrr::map(names(seurat_list), function(x) {
  p <- seurat_list[[x]]@meta.data %>%
    ggplot(aes(time_point, nFeature_RNA, fill = subproject)) +
      geom_violin() +
      scale_y_log10() +
      labs(title = x, x = "", y = "Number of Detected Genes") +
      theme_bw() +
      theme(
        plot.title = element_text(hjust = 0.5, size = 13),
        axis.text = element_text(size = 11),
        axis.title = element_text(size = 13),
        legend.title = element_blank()
      ) 
  p
})
names(n_features_ggs) <- names(seurat_list)
n_features_ggs
```



# Filter out cells

We aim to detect and exclude empty droplets or lysed cells. Lysed cells have 3 hallmarks: (1) low library size (total UMI), (2) low library complexity (number of detected genes) and (3) high fraction of mitochondrial expression (cytosolic mRNA leaks out of the cell). Let us start by visualizing their univariate distributions.


## Library Size

```{r fig.wide=TRUE}
lib_size_hist <- seurat@meta.data %>%
  plot_histogram_qc(x = "nCount_RNA", x_lab = "Library Size (log10(total UMI))") +
  geom_vline(xintercept = min_lib_size, linetype = "dashed", color = "red")
lib_size_hist1 <- lib_size_hist +
    scale_x_log10()
lib_size_hist2 <- lib_size_hist +
    scale_x_continuous(limits = c(0, 4000)) +
    xlab("Library Size (total UMI)") +
    theme_pubr()
lib_size_hist1 + lib_size_hist2
```


## Number of detected genes

```{r fig.wide=TRUE}
n_genes_hist1 <- seurat@meta.data %>%
  plot_histogram_qc(x = "nFeature_RNA", x_lab = "Number of Detected Genes") +
  geom_vline(xintercept = min_n_genes, linetype = "dashed", color = "red")
n_genes_hist2 <- n_genes_hist1 +
  scale_x_continuous(limits = c(0, 2000)) 
n_genes_hist1 + n_genes_hist2
```


## Fraction of mitochondrial expression

```{r}
# Calculate percentage of mitochondrial and ribosomal expression
seurat$pct_mt <- PercentageFeatureSet(seurat, pattern = "^MT-")


# Plot
pct_mt_hist <- seurat@meta.data %>%
  plot_histogram_qc(x = "pct_mt", x_lab = "% Mitochondrial Expression") +
  geom_vline(xintercept = max_pct_mt, linetype = "dashed", color = "red") +
  scale_x_continuous(limits = c(0, 100))
pct_mt_hist
```


## Fraction of ribosomal expression

Let us also calculate and visualize the proportion of ribosomal expression:

```{r}
seurat$pct_ribosomal <- PercentageFeatureSet(
  seurat,
  pattern = "^RPS"
)
pct_ribosomal_hist <- seurat@meta.data %>%
  plot_histogram_qc(x = "pct_ribosomal", x_lab = "% Ribosomal Expression")
pct_ribosomal_hist
```


## Joint QC metrics

It is important to assess how these variables covary, since metabolically active cells might also have a high mitochondrial expression:

```{r}
# number of detected genes VS library size
n_genes_vs_lib_size <- FeatureScatter(
  seurat,
  feature1 = "nCount_RNA",
  feature2 = "nFeature_RNA",
  pt.size = 0.15,
  cols = rep("black", length(levels(Idents(seurat))))
)
n_genes_vs_lib_size <- n_genes_vs_lib_size +
  labs(x = "Library Size (total UMI)", y = "Number of Detected Genes") +
  theme(legend.position = "none", plot.title = element_blank())
n_genes_vs_lib_size +
  geom_vline(xintercept = min_lib_size, linetype = "dashed", color = "red") +
  geom_hline(yintercept = min_n_genes, linetype = "dashed", color = "red")

  
# % mitochondrial expression VS library size
pct_mt_vs_lib_size <- FeatureScatter(
  seurat,
  feature1 = "nCount_RNA",
  feature2 = "pct_mt",
  pt.size = 0.15,
  cols = rep("black", length(levels(Idents(seurat))))
)
pct_mt_vs_lib_size <- pct_mt_vs_lib_size +
  scale_x_log10()+
  labs(x = "Library Size (total UMI)", y = "% Mitochondrial Expression") +
  theme(legend.position = "none", plot.title = element_blank())
pct_mt_vs_lib_size +
  geom_vline(xintercept = min_lib_size, linetype = "dashed", color = "red") +
  geom_hline(yintercept = max_pct_mt, linetype = "dashed", color = "red")
```


## Subset empty droplets and lysed cells

```{r}
metadata_before_qc <- seurat@meta.data
is_low_quality <- 
  seurat$nCount_RNA < min_lib_size |
  seurat$nFeature_RNA < min_n_genes |
  seurat$pct_mt > max_pct_mt
table(is_low_quality)
seurat$keep_cells <- !is_low_quality
Idents(seurat) <- "keep_cells"
seurat
seurat <- subset(seurat, idents = TRUE)
seurat
metadata_after_qc <- seurat@meta.data
```


## QC summary table

```{r}
qc_before <- metadata_before_qc %>%
  group_by(donor_id, sample_id, time_point) %>% 
  summarise(num_cells_before_qc = n())
qc_after <- metadata_after_qc %>%
  group_by(donor_id, sample_id, time_point) %>%
  summarise(
    num_cells_after_qc = n(),
    average_library_size = mean(nCount_RNA),
    average_num_detected_genes = mean(nFeature_RNA),
    average_mitochondrial_fraction = mean(pct_mt)
  )
qc_table <- left_join(qc_before, qc_after, by = "sample_id")
DT::datatable(qc_table)
```


# Filter out genes

According to [Luecken MD et al.](https://www.embopress.org/doi/10.15252/msb.20188746): "A guideline to setting this threshold is to use the minimum cell cluster size that is of interest and leaving some leeway for dropout effects. For example, filtering out genes expressed in fewer than 20 cells may make it difficult to detect cell clusters with fewer than 20 cells. For datasets with high dropout rates, this threshold may also complicate the detection of larger clusters. The choice of threshold should scale with the number of cells in the dataset and the intended downstream analysis."

Since we want to detect rare cell types in the tonsil, we will be very permissive and retain genes that are expressed in at least `r min_cells` cells.


## Set threshold

```{r fig.wide=TRUE}
n_cells <- Matrix::rowSums(seurat[["RNA"]]@counts > 0)
gene_qc <- n_cells %>% 
  as.data.frame() %>% 
  ggplot(aes(n_cells)) + 
    geom_histogram(bins = 100, alpha = 0.75) +
    scale_x_log10("Number of cells") +
    theme_bw() 
gene_qc +
  geom_vline(xintercept = min_cells, linetype = "dashed", color = "red")
```


## Plot genes with highest expression

```{r fig.height=10}
top_50_genes <- sort(n_cells, decreasing = TRUE)[1:50]
top_50_genes_df <- data.frame(
  gene = names(top_50_genes),
  n_cells = top_50_genes
)
top_50_genes_df %>%
  ggplot(aes(fct_reorder(gene, n_cells), n_cells)) +
    geom_point() +
    labs(x = "", y = "Number of expressing cells") +
    coord_flip()
```


## Filter

```{r}
kept_genes <- rownames(seurat)[n_cells > min_cells]
table(n_cells > min_cells)
seurat
seurat <- subset(seurat, features = kept_genes)
seurat
```


# Save filtered Seurat object

```{r}
saveRDS(seurat, path_to_save)
```
    

# Session Information

```{r}
sessionInfo()
```