---
title: 'Dimensionality Reduction (scATAC-seq)'
author: "Ramon Massoni-Badosa"
date: "2021/04/09"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```


# Introduction

Here we will run the standard [Signac pipeline](https://satijalab.org/signac/articles/pbmc_vignette.html) that will allow us to cluster cells and visualize them in UMAP space.


# Pre-processing

## Load packages

```{r}
library(Seurat)
library(Signac)
library(EnsDb.Hsapiens.v86)
library(GenomicRanges)
library(GEOquery)
library(tidyverse)
library(glue)
```


## Define parameters

```{r}
path_to_obj <- here::here("results/R_objects/Penter2021/scATAC-seq/1.seurat_obj_init_Penter2021_atac.rds")
path_to_save <- here::here("results/R_objects/Penter2021/scATAC-seq/2.seurat_obj_dimred_Penter2021_atac.rds")


# Thresholds
min_tss_enrichment <- 3.25
min_num_frag <- 1000
```


## Read data

```{r}
seurat_obj <- readRDS(path_to_obj)
```


# QC

## Calculate QC metrics

```{r}
seurat_obj <- NucleosomeSignal(seurat_obj)
seurat_obj <- TSSEnrichment(seurat_obj, fast = FALSE)
```


## Plot

```{r}
# Transcription start site enrichment plot
TSSPlot(seurat_obj)
VlnPlot(seurat_obj, "TSS.enrichment", pt.size = 0) +
  NoLegend() +
  geom_hline(yintercept = min_tss_enrichment, color = "red", linetype = "dashed") +
  theme(axis.title.x = element_blank())


# Number of fragments per cell
VlnPlot(seurat_obj, "nFeature_ATAC", pt.size = 0) +
  NoLegend() +
  geom_hline(yintercept = min_num_frag, color = "red", linetype = "dashed") +
  scale_y_log10() +
  theme(axis.title.x = element_blank())
```


## Filter

According to [Penter et al.](https://cancerdiscovery.aacrjournals.org/content/11/12/3048): "The ArchR manual was followed for exclusion of low- quality cells (transcription start site (TSS) enrichment <4 and fragments per cell <1000)."

Looking at the plots above, a TSS enrichment of 4 seems too stringent, so we will lower it to `r min_tss_enrichment`.

```{r}
is_low_quality <- with(seurat_obj@meta.data,
  TSS.enrichment < min_tss_enrichment |
  nFeature_ATAC < min_num_frag
)
table(is_low_quality)
seurat_obj <- subset(seurat_obj, cells = colnames(seurat_obj)[!is_low_quality])
```


# Dimensionality reduction

```{r}
seurat_obj <- RunTFIDF(seurat_obj)
seurat_obj <- FindTopFeatures(seurat_obj, min.cutoff = "q25")
seurat_obj <- RunSVD(seurat_obj)
DepthCor(seurat_obj)
seurat_obj <- RunUMAP(object = seurat_obj, reduction = "lsi", dims = 2:30)
seurat_obj <- FindNeighbors(object = seurat_obj, reduction = "lsi", dims = 2:30)


# Plot
DimPlot(seurat_obj, group.by = "disease_state")
DimPlot(seurat_obj, group.by = "dataset_id")
DimPlot(seurat_obj, group.by = "sample_description")
DimPlot(seurat_obj, group.by = "tissue")
```


# Save

```{r}
saveRDS(seurat_obj, path_to_save)
```


# Session Information

```{r}
sessionInfo()
```

