---
title: 'Patient-specific differential expression analysis (RT vs CLL)'
author: "Ramon Massoni-Badosa"
date: "2021/04/09"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```


# Introduction

The purpose of this notebook is to perform a patient-specific differential expression analysis (DEA) between the RT and CLL clusters. We will focus on the quiescent RT populations, since otherwise the proliferation signature will dominate the analysis.


# Pre-processing

## Load packages

```{r}
library(Seurat)
library(SeuratWrappers)
library(harmony)
library(UpSetR)
library(ggrepel)
library(ggforce)
library(tidyverse)
```


## Define parameters

```{r}
# Paths
path_to_12 <- here::here("results/R_objects/6.seurat_annotated_12.rds")
path_to_19 <- here::here("results/R_objects/6.seurat_annotated_19.rds")
path_to_63 <- here::here("results/R_objects/patient_63/3.seurat_annotated.rds")
path_to_365 <- here::here("results/R_objects/6.seurat_annotated_365.rds")
path_to_3299 <- here::here("results/R_objects/6.seurat_annotated_3299.rds")
path_to_save_rds <- here::here("6-differential_expression_analysis/tmp/patient_specific_differential_expression_analysis_rt_vs_cll.rds")
path_to_save_xlsx <- here::here("results/tables/DEA/patient_specific_differential_expression_analysis_rt_vs_cll.xlsx")


# Colors
color_palette <- c("black", "gray", "red", "yellow", "violet", "green4",
                   "blue", "mediumorchid2", "coral2", "blueviolet",
                   "indianred4", "deepskyblue1", "dimgray", "deeppink1",
                   "green", "lightgray", "hotpink1")


# Source functions
source(here::here("bin/utils.R"))


# Thresholds
alpha <- 0.05
min_avg_log2FC <- 0.25
selected_avg_log2FC_l <- c(
  "12" = 1.25,
  "19" = 0.75,
  "3299" = 0.75,
  "365" = 1.5,
  "63" = 1.25  
)
selected_avg_log2FC <- 1.25
selected_pct_cells <- 10
selected_significance_alpha <- alpha
```


## Load data

```{r}
paths_to_load <- c(
  "12" = path_to_12,
  "19" = path_to_19,
  "63" = path_to_63,
  "365" = path_to_365,
  "3299" = path_to_3299
)
seurat_list <- purrr::map(paths_to_load, readRDS)
seurat_list


# Plot
(umaps_annotations <- purrr::map(names(seurat_list), function(x) {
  p <- plot_annotation(
    seurat_list[[x]],
    pt_size = 0.5,
    colors_reference = color_annotations,
    patient_id = x
  )
  p +
    ggtitle(x) +
    theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
}))
```


# Differential expression analysis

```{r}
comparisons <- list(
  "12" = list(
    richter = c("CCND2lo RT", "CCND2hi RT"),
    cll = c("CXCR4hiCD27lo", "CXCR4loCD27hi", "MIR155HGhi", "MZB1hiIGHMhiXBP1hi")
  ),
  "19" = list(
    richter = c("MIR155HGhi RT", "TP53INP1hi RT"),
    cll = c("CXCR4hiCD27lo", "CXCR4loCD27hi", "MIR155HGhi CLL")
  ),
  "63" = list(
    richter = "RT quiescent",
    cll = "CLL"
  ),
  "365" = list(
    richter = c("RT quiescent", "CXCR4loCD27hi RT", "MIR155HGhi RT"),
    cll = c("CLL")
  ),
  "3299" = list(
    richter = "RT",
    cll = c("CXCR4hiCD27lo", "CXCR4loCD27hi", "CD83loMIR155HGhi", "CD83hiMIR155HGhi")
  )
)
dea <- purrr::map(names(comparisons), function(x) {
  print(x)
  seurat_obj <- seurat_list[[x]]
  Idents(seurat_obj) <- "annotation_final"
  df <- FindMarkers(
    seurat_obj,
    ident.1 = comparisons[[x]][["richter"]],
    ident.2 = comparisons[[x]][["cll"]],
    only.pos = FALSE,
    logfc.threshold = 0
  )
  df <- df %>%
    rownames_to_column("gene") %>%
    arrange(desc(avg_log2FC))
  df
}) 
names(dea) <- names(comparisons)
dea <- purrr::map(names(dea), function(x) {
  df <- dea[[x]]
  df$direction <- case_when(
    df$p_val_adj < alpha & df$avg_log2FC > min_avg_log2FC ~ "up",
    df$p_val_adj < alpha & df$avg_log2FC < 0 & abs(df$avg_log2FC) > min_avg_log2FC ~ "down"
  )
  df$direction[is.na(df$direction)] <- "not sig."
  pct_cells <- rowMeans(seurat_list[[x]][["RNA"]]@counts > 0) * 100
  df$pct_cells_expressing <- pct_cells[df$gene]
  df
})
names(dea) <- names(comparisons)
upregulated_richter <- purrr::map(dea, function(df) df$gene[df$direction == "up"])
downregulated_richter <- purrr::map(dea, function(df) df$gene[df$direction == "down"])
print("DEA patient 12")
DT::datatable(dea$`12`, options = list(scrollX = TRUE))
print("DEA patient 19")
DT::datatable(dea$`19`, options = list(scrollX = TRUE))
print("DEA patient 63")
DT::datatable(dea$`63`, options = list(scrollX = TRUE))
print("DEA patient 365")
DT::datatable(dea$`365`, options = list(scrollX = TRUE))
print("DEA patient 3299")
DT::datatable(dea$`3299`, options = list(scrollX = TRUE))
```


# Upset plots

```{r}
upset_upregulated <- upset(fromList(upregulated_richter), order.by = "freq")
upset_downregulated <- upset(fromList(downregulated_richter), order.by = "freq")
upset_upregulated
upset_downregulated
```


# MA plots

```{r fig.wide=TRUE}
ma_plots <- purrr::map(names(dea), function(x) {
  df <- dea[[x]]
  selected_avg_log2FC <- selected_avg_log2FC_l[x]
  p <- ma_plot(
    df,
    selected_avg_log2FC = selected_avg_log2FC,
    selected_pct_cells = selected_pct_cells,
    selected_significance_alpha = selected_significance_alpha
  ) +
    ggtitle(x) +
    theme(plot.title = element_text(hjust = 0.5))
  p
})
names(ma_plots) <- names(dea)
ma_plots
```


# Save

```{r}
saveRDS(dea, path_to_save_rds)
openxlsx::write.xlsx(dea, path_to_save_xlsx)
```


# Session Information

```{r}
sessionInfo()
```


